---
title: "Tutorial"
output:
  html_document:
    theme: united
    df_print: kable
    toc: true
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`Gretool` package is implemented in R and publicly available at <https://github.com/thereallda/Gretool>.

## Setup

For this tutorial, we will demonstrate the `Gretool` workflow by using a glyco-capped RNA-seq data from human.

We start by reading in the data, including count matrix and metadata. The metadata requires at lease two columns:

1\. The *condition* column specifies the biological group of each sample, such as **A.Input**, indicating that the sample is an input sample from biological condition A.

2\. The *enrich* column denotes the enrichment group of each sample, such as **Input**, representing the input library.

```{r,warning=FALSE,message=FALSE}
library(Gretool)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)

# read in metadata and counts matrix
counts_mat <- read.csv("data/Counts.csv", row.names = 1)
meta <- read.csv("data/metadata.csv", comment.char = "#")
head(meta)
# rownames of metadata should be consistent with the colnames of counts_mat
rownames(meta) <- meta$id
```

Also, other information, such as batch group can be included in the metadata.

Next, we use the count matrix and metadata to create a `Gretool` object.

The `Gretool` object serves as a container that contains both data (raw and normalized counts data) and analysis (e.g., normalization performance evaluation score and enrichment results).

When constructing `Gretool` object, we can provide following arguments:

-   The id of input (`input.id = "Input"`) and enrichment (`enrich.id = "Enrich"`), same as the enrich column in the metadata (**required**).

-   Prefix of spike-in genes (`spike.in.prefix = "^FB"`) (optional);

-   The id of synthetic spike-in (`synthetic.id = c("Syn1", "Syn2")`), same with row names in `counts_mat` (optional);

```{r}
# create object
Gretool <- createGretool(data = counts_mat,
                         col.data = meta,
                         input.id = "Input",
                         enrich.id = "Enrich"
                         )
Gretool
```

## Quick start

`Gretool` starts with quality control step to filter lowly-expressed genes. This step is performed with `FilterLowExprGene` by keeping genes with at least `min.count` in at least n samples. n is determined by the smallest group sample size specifying in `group` .

```{r}
Gretool <- FilterLowExprGene(Gretool, group = Gretool$condition, min.count = 10)
Gretool
```
### Normalization and assessment

The main `Gretool` function can be used to perform gene selection, normalization, and evaluation. This function return selected gene set in `rowData`, normalized count matrix in `counts` slot (when `return.norm=TRUE`), normalization factors (`norm_factors` slot), evaluation metrics and scores in `norm_metrics` and `norm_scores` slots, respectively. Detailed of these steps are described bellow.

```{r}
Gretool <- Gretool(Gretool, 
               scaling.method = c( "TMM", "DESeq"),
               ruv.norm = TRUE, ruv.k = 3,
               eval.pam.k = 2:3, eval.pc.n = 3, 
               return.norm = TRUE
               )
```

## Effect of normalization

Here, we use the top-ranked procedure DESeq_RUVs_k3 for downstream analysis.

We perform PCA based on the count matrix from sample of interest before and after the normalization, for demonstrating the effect of normalization.

```{r, fig.width=12, fig.height=6}
# get scores for normalization
norm_scores <- getScore(Gretool)

head(norm_scores)

# select normalization
norm_method <- rownames(norm_scores[1,])

# get normalized counts
norm_data <- Gretool::Counts(Gretool, slot = "sample", method = norm_method)

# get normalization factors
norm_factors <- getFactor(Gretool, slot = "sample", method = norm_method)

# create sample name, e.g., A.Input.1
samples_name <- paste(Gretool$condition, Gretool$replicate, sep=".")

# PCA for raw count
p1 <- PCAplot(Gretool::Counts(Gretool, slot="sample", "Raw"), 
            color = Gretool$condition,
            shape = Gretool$enrich,
            label = samples_name, 
            vst.norm = TRUE) + 
  ggtitle("Raw")

# PCA for normalized count
p2 <- PCAplot(log1p(norm_data), 
            color = Gretool$condition,
            shape = Gretool$enrich,
            label = samples_name, 
            vst.norm = FALSE) + 
  ggtitle("After normalization")

# combine two plots
p1 + p2
```

## Find enrichment

`Gretool` package can help you find enrichment genes from each biological groups via differential expression. It can identify genes that significantly increased in enrichment samples compared to input samples.

`FindEnrichment` automates this processes for all groups provided in `condition` column.

By default, enriched genes are defined as fold change of normalized transcript counts ≥ 2 (`logfc.cutoff = 1`), FDR \< 0.05 (`p.cutoff = 0.05`) in enrichment samples compared to those in input samples.

Use `getEnrichment` to retrieve a list of enrichment result tables.

```{r}
# find all enriched genes
Gretool <- FindEnrichment(Gretool, 
                          slot="sample", 
                          norm.method = norm_method, 
                          logfc.cutoff = 1, 
                          p.cutoff = 0.05)

# get filtered enrichment results
res.sig.ls <- getEnrichment(Gretool, slot="sample", filter=TRUE)
names(res.sig.ls) <- c('H','T')
# count number of enrichment in each group
unlist(lapply(res.sig.ls, nrow))
```

Each enrichment table is a `data.frame` with a list of genes as rows, and associated information as columns (GeneID, logFC, p-values, etc.). The following columns are present in the table:

-   `GeneID`: ID of genes.
-   `logFC`: log2 fold-change between enrichment and input samples. Positive values indicate that the gene is more highly enriched in the enrichment group.
-   `logCPM`: log2 CPM (counts per million) of the average expression of all samples.
-   `LR`: Likelihood ratio of the likelihood ratio test.
-   `PValue`: p-value from the likelihood ratio test.
-   `FDR`: False discovery rate of the p-value, default "BH" method is applied.

## Differential modification analysis

You can perform differential modification analysis with `DiffModification`

```{r}
# create biological column in metadata
meta$bio <- gsub('\\..*','',meta$condition)
meta

# testing 
diff_mod <- DiffModification(
  Gretool,
  norm.method = norm_method,
  col.data = meta,
  contrast.df = data.frame(Group1=c('H'), 
                           Group2=c('T')),
  assay.colname = 'enrich',
  biology.colname = 'bio',
  input.id = 'Input',
  enrich.id = 'Enrich',
  eps = 1e-4,
  shrink_lambda = 0
)

head(diff_mod[[1]])
```
By default, `DiffModification` output all differential modification results for all genes. You can subset the table based on the enriched genes from `FindEnrichment`

```{r}
enrich_ids <- unique(unlist(lapply(res.sig.ls, function(x) x$GeneID)))

diff_mod_enrich <- subset(diff_mod[[1]], GeneID %in% enrich_ids)
dim(diff_mod_enrich)
```

You can visualize the differential results with scatter plot 
```{r}
typedf1 <- data.frame(table(diff_mod_enrich$mod.type)) %>%
  mutate(label = paste0(Var1,' (n=',Freq,')'),
         prop = round(100*Freq/sum(Freq),1))

ggplot(diff_mod_enrich, aes(T.logFC,H.logFC, color=mod.type)) +
  geom_point(aes(size = -log10(mod.FDR)), alpha=0.8, shape=16, stroke=NA) +
  theme_classic() +
  theme(legend.position = 'right',
        axis.text = element_text(color='black'),
        plot.title = element_text(hjust=0.5))+
  scale_color_manual(values=c("#A83C1E","#4F99B4","grey80"),
                     labels=typedf1$label) +
  labs(color='', 
       title='H vs T')
```

## Session Info

<details>

<summary>**Session Info**</summary>

```{r}
sessionInfo()
```

</details>
